---
title: "Notebook to reproduce manuscript figures"
params:
    spatial_features_dir: "data/features_tcga_skcm_ff"
    tcga_bagaev_subtypes: "data/bagaev_file.xlsx"
    tcga_survival: "data/tcga_survival_file.xlsx"
    bootstrap_model: "data/bootstrap_model.RData"
output: html_document
---
  <style type="text/css">
  .smaller {
    font-size: 10px
    font-style: normal
  }
</style>


```{r setup, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# tcga_bagaev_subtypes: "./data/mmc6.xlsx",  tcga_survival: "./data/mmc1.xlsx"

params <- list()
params$spatial_features_dir <- "data/features_tcga_skcm_ff"
params$tcga_bagaev_subtypes <- "data/bagaev_file.xlsx"
params$tcga_survival <- "data/tcga_survival_file.xlsx"
params$bootstrap_model <- "data/bootstrap_model.RData"
# packages
suppressPackageStartupMessages({
  library("dplyr")
  library("corrplot")
  library("gplots")
  library("RColorBrewer")
  library("Matrix")
  library("ggplot2")
  library("ggforce")
  library("ggpubr")
  library("reshape2")
  library("ggrepel")
  library("tidyr")
  library("plyr")
  library("purrr")
  library("gridExtra")
  library("rstatix")
  library("ComplexHeatmap")
  library("survival")
  library("ggfortify")
  library("survminer")
})

colors_mfp <- c("#64a860", "#9970c1", "#cc545e", "#b98d3e")
names(colors_mfp) <- c("D", "F", "IE/F", "IE")

# change default settings to fit plots
knitr::opts_chunk$set(fig.height = 8, fig.width = 14)

# Read TCGA-SKCM features ------------------------------------------------------

tcga_mat <- data.table::fread(file = "./data/spatial_features_matrix_TCGA.csv")
tcga_mat <- tcga_mat %>% tibble::column_to_rownames(var = "V1")

# Read CPTAC-CM features ------------------------------------------------------

cptac_mat <- data.table::fread(file = "./data/spatial_features_matrix_CPTAC.csv")
cptac_mat <- cptac_mat %>% tibble::column_to_rownames(var = "V1")

# Read Bagaev subtypes for TCGA data -------------------------------------------
bagaev_mfp <- readxl::read_xlsx(path = params$tcga_bagaev_subtypes, sheet = 1, skip = 1)
bagaev_mfp <- bagaev_mfp %>% filter(Cohort == "TCGA-SKCM")
colnames(bagaev_mfp)[1] <- "TCGA_patient_ID"

# Read Survival data for TCGA data -------------------------------------------
survival_tcga_liu <- readxl::read_xlsx(path = params$tcga_survival, sheet = 1)
```

## Figure 1B

Cell type-specific probability maps for four examples of melanoma slides from distinct microenvironments: immune-depleted (D), fibrotic (F), immune-enriched fibrotic (IE/F) and immune-enriched (IE) subtypes. 

** Note that the full slide is shown here. **

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 22, fig.height = 22}
##  Read cell type probabilities
cell_prob <- data.table::fread(
  file = file.path(
    params$spatial_features_dir,
    "tcga_validation_tile_predictions_proba.csv"
  ),
  sep = "\t"
)

## Pivot to long df
cell_prob <- cell_prob %>%
  select(slide_submitter_id, CAFs, T_cells, endothelial_cells, tumor_purity, Coord_X, Coord_Y) %>%
  pivot_longer(
    cols = !c(slide_submitter_id, Coord_X, Coord_Y),
    names_to = "celltype", values_to = "prob"
  )

##  Add bagaev subtypes data
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
cell_prob$TCGA_patient_ID <- substr(cell_prob$slide_submitter_id, 1, 12)
cell_prob <- inner_join(cell_prob, tmp, by = "TCGA_patient_ID")

##  Desert
D_example <- "TCGA-YG-AA3P"
# TCGA-Z2-AA3S
# TCGA-D3-A8GL

##  Fibrotic
F_example <- "TCGA-EE-A3JB"
# TCGA-D3-A2JA
#  TCGA-FR-A2OS

##  Immune-enrinched fibrotic
IEF_example <- "TCGA-EE-A2ME" # IE/F
# TCGA-FS-A1ZH
# TCGA-D3-A1Q3

##  Immune-enrinched
IE_example <- "TCGA-ER-A1A1"
# TCGA-D3-A8GM
# TCGA-WE-A8K4

##  Visualize selected slides ##
examples <- c(D_example, F_example, IEF_example, IE_example)

gg_maps <- lapply(examples, function(X) {
  example <- cell_prob %>% filter(TCGA_patient_ID %in% X)

  # split data for each "facet"
  df <- split(example, f = example$celltype)

  plot_CAFs <- df$CAFs %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Greens", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_endo <- df$endothelial_cells %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Purples", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_tcells <- df$T_cells %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Blues", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_tumor <- df$tumor_purity %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  if (X == examples[4]) {
    ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor,
      nrow = 4, ncol = 1, align = "v"
    )
  } else {
    ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor,
      nrow = 4, ncol = 1, align = "v", legend = "none"
    )
  }
})

ggarrange(
  plot_list = gg_maps,
  nrow = 1, ncol = 4, align = "h", common.legend = TRUE
)
```

## Figure 2

Slide-level quantification of each cell type
Co-localization assessment of cell type pairs across the distinct microenvironment subtypes 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 12}
# Cell type probability -------------------------------------------

## Calculate median probability for each slide
cell_prob_mean <- cell_prob %>%
  dplyr::group_by(TCGA_patient_ID, celltype, MFP) %>%
  dplyr::summarise(mean_prob = mean(prob))

cell_prob_mean$MFP <- factor(cell_prob_mean$MFP, levels = names(colors_mfp))
cell_prob_mean <- cell_prob_mean %>% ungroup()

##  Pairwise statistical test #
stat.test <- cell_prob_mean %>%
  group_by(celltype) %>%
  wilcox_test(mean_prob ~ MFP) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
  add_xy_position(x = "celltype", group = "MFP") %>%
  filter(!p.adj.signif == "ns")

##  Add factor levels
cell_prob_mean$celltype <- factor(cell_prob_mean$celltype,
  levels = c(
    "CAFs", "endothelial_cells",
    "T_cells", "tumor_purity"
  )
)

## Plot
bxp_prob <- ggpubr::ggboxplot(
  cell_prob_mean,
  x = "celltype", y = "mean_prob",
  fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
  add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
  scale_x_discrete(labels = c("CAFs", "Endo cells", "T cells", "Tumor cells")) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 16, color = "black"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16, color = "black"),
    legend.position = "top", legend.direction = "horizontal",
    legend.title = element_text(size = 16, color = "black"),
    legend.text = element_text(size = 16, color = "black")
  ) +
  labs(y = "Mean probability")

fig2a <- bxp_prob +
  stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

# Cell type co-localization -------------------------------------------

cell_coloc <- data.table::fread(
  file = file.path(params$spatial_features_dir, "FF_features_coloc_fraction.csv"),
  sep = "\t"
)

##  Change cell-type names
cell_coloc <- cell_coloc %>%
  mutate(pair = gsub("T_cells", "Tcell", pair))
cell_coloc <- cell_coloc %>%
  mutate(pair = gsub("endothelial_cells", "Endo", pair))
cell_coloc <- cell_coloc %>%
  mutate(pair = gsub("tumor_purity", "Tumor", pair))
cell_coloc <- cell_coloc %>%
  mutate(pair = gsub("CAFs", "CAF", pair))

##  Add Bagaev subtypes
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
cell_coloc$TCGA_patient_ID <- substr(cell_coloc$slide_submitter_id, 1, 12)
cell_coloc <- inner_join(cell_coloc, tmp, by = "TCGA_patient_ID")
cell_coloc$MFP <- factor(cell_coloc$MFP, levels = names(colors_mfp))

cell_coloc$pair <- factor(cell_coloc$pair, levels = c(
  "CAF-Endo", "CAF-Tcell", "CAF-Tumor",
  "Endo-Tumor", "Tcell-Endo", "Tcell-Tumor"
))
##  Pairwise statistical test
stat.test <- cell_coloc %>%
  group_by(pair) %>%
  wilcox_test(frac ~ MFP) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
  add_xy_position(x = "pair", group = "MFP") %>%
  filter(!p.adj.signif == "ns")

##  Add factor levels
cell_coloc$pair <- factor(cell_coloc$pair, levels = unique(test_subtypes$pair))

##  Plot
bxp_coloc <- ggpubr::ggboxplot(
  cell_coloc,
  x = "pair", y = "frac",
  fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
  add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16, color = "black"),
    legend.position = "top", legend.direction = "horizontal",
    legend.title = element_text(size = 16, color = "black"),
    legend.text = element_text(size = 16, color = "black")
  ) +
  labs(y = "Co-localization")

fig2b <- bxp_coloc +
  stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

ggarrange(fig2a,
  fig2b,
  nrow = 1, ncol = 2, align = "h", common.legend = TRUE
)
```

## Figure 4

Number of shortest paths with maximum path length equal to two (SP2).
Size of the largest connected component (LCC) 
Proximity between cell type-specific clusters classified as highly or lowly abundant.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 12}
# Cell type SP2 -------------------------------------------

SP2 <- data.table::fread(
  file = file.path(params$spatial_features_dir, "FF_features_shortest_paths_thresholded.csv"),
  sep = "\t"
)

SP2 <- SP2 %>%
  dplyr::mutate(TCGA_patient_ID = substr(slide_submitter_id, 1, 12)) %>%
  dplyr::rename("celltype_pair" = "pair")

#  Change cell-type names
SP2 <- SP2 %>%
  mutate(celltype_pair = gsub("T_cells", "Tcell", celltype_pair))
SP2 <- SP2 %>%
  mutate(celltype_pair = gsub("endothelial_cells", "Endo", celltype_pair))
SP2 <- SP2 %>%
  mutate(celltype_pair = gsub("tumor_purity", "Tumor", celltype_pair))
SP2 <- SP2 %>%
  mutate(celltype_pair = gsub("CAFs", "CAF", celltype_pair))

#  Add MFP data
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
SP2$TCGA_patient_ID <- substr(SP2$slide_submitter_id, 1, 12)
SP2 <- inner_join(SP2, tmp, by = "TCGA_patient_ID")
SP2$MFP <- factor(SP2$MFP, levels = c("D", "F", "IE/F", "IE"))

##  Pairwise statistical test
stat.test <- SP2 %>%
  group_by(celltype_pair) %>%
  wilcox_test(n_paths ~ MFP) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
  add_xy_position(x = "celltype_pair", group = "MFP") %>%
  filter(!p.adj.signif == "ns")

SP2$celltype_pair <- factor(SP2$celltype_pair,
  levels = unique(test_subtypes$celltype_pair)
)

## Plot
bxpSP2 <- ggpubr::ggboxplot(
  SP2,
  x = "celltype_pair", y = "n_paths",
  fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
  add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 16, color = "black"),
    axis.text.x = element_text(size = 16, color = "black", angle = 45, vjust = 1, hjust = 1),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16, color = "black"),
    legend.position = "top", legend.direction = "horizontal"
  ) +
  labs(y = "SP2")

Fig4a <- bxpSP2 +
  stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

# Cell type LCC -------------------------------------------

LCC <- data.table::fread(
  file = file.path(params$spatial_features_dir, "FF_features_lcc_fraction.csv"),
  sep = "\t"
)

## Pivot to longer
LCC <- LCC %>%
  pivot_longer(cols = !c("slide_submitter_id"), names_to = "celltype", values_to = "LCC")

##  Change cell-type names
LCC <- LCC %>%
  mutate(celltype = gsub(".clusters", "", celltype))
LCC <- LCC %>%
  mutate(celltype = gsub("T.cells", "Tcell", celltype))
LCC <- LCC %>%
  mutate(celltype = gsub("endothelial.cells", "Endo", celltype))
LCC <- LCC %>%
  mutate(celltype = gsub("tumor.purity", "Tumor", celltype))
LCC <- LCC %>%
  mutate(celltype = gsub("CAFs", "CAF", celltype))
LCC <- LCC %>%
  mutate(celltype = gsub("LCC.", "", celltype))

##  Add Bagev subtypes
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
LCC$TCGA_patient_ID <- substr(LCC$slide_submitter_id, 1, 12)
LCC <- inner_join(LCC, tmp, by = "TCGA_patient_ID")
LCC$MFP <- factor(LCC$MFP, levels = c("D", "F", "IE/F", "IE"))

## Factor levels
LCC$celltype <- factor(LCC$celltype, levels = c("CAF", "Endo", "Tcell", "Tumor"))

##  Pairwise statistical test
stat.test <- LCC %>%
  group_by(celltype) %>%
  wilcox_test(LCC ~ MFP) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
  add_xy_position(x = "celltype", group = "MFP") %>%
  filter(!p.adj.signif == "ns")

LCC$celltype <- factor(LCC$celltype, levels = unique(test_subtypes$celltype))

# Create a box plot
bxp_LCC <- ggpubr::ggboxplot(
  LCC,
  x = "celltype", y = "LCC",
  fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
  add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
  theme_minimal() +
  scale_x_discrete(labels = c("CAFs", "Endo cells", "T cells", "Tumor cells")) +
  theme(
    axis.text.y = element_text(size = 16, color = "black"),
    axis.text.x = element_text(size = 16, color = "black", angle = 0),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16, color = "black"),
    legend.position = "top", legend.direction = "horizontal"
  ) +
  labs(y = "LCC")

Fig4b <- bxp_LCC +
  stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

# Cell type Proximity of clusters -------------------------------------------

prox_clust <- data.table::fread(
  file = file.path(params$spatial_features_dir, "FF_features_clust_indiv_schc_prox.csv"),
  sep = "\t"
)

##  Add Bagaev subtypes
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
prox_clust$TCGA_patient_ID <- substr(prox_clust$slide_submitter_id, 1, 12)
prox_clust <- inner_join(prox_clust, tmp, by = "TCGA_patient_ID")

prox_clust <- prox_clust %>%
  dplyr::rename(pair = `pair (comparison)`) %>%
  dplyr::select(slide_submitter_id, pair, proximity, TCGA_patient_ID, MFP)

prox_clust <- prox_clust %>%
  mutate(pair = gsub("tumor_purity", "Tumor", pair))
prox_clust <- prox_clust %>%
  mutate(pair = gsub("CAFs", "CAF", pair))
prox_clust <- prox_clust %>%
  mutate(pair = gsub("endothelial_cells", "Endo", pair))
prox_clust <- prox_clust %>%
  mutate(pair = gsub("T_cells", "Tcell", pair))

prox_clust <- prox_clust[-grep("(low-low)", prox_clust$pair), ]
prox_clust$pair <- gsub("high", "H", prox_clust$pair)
prox_clust$pair <- gsub("low", "L", prox_clust$pair)

newlevels <- c(
  "CAF-CAF (L-H)", "CAF-CAF (H-H)",
  "CAF-Endo (L-H)", "CAF-Endo (H-H)", "CAF-Endo (H-L)",
  "CAF-Tcell (L-H)", "CAF-Tcell (H-H)", "CAF-Tcell (H-L)",
  "CAF-Tumor (L-H)", "CAF-Tumor (H-H)", "CAF-Tumor (H-L)",
  "Endo-Endo (L-H)", "Endo-Endo (H-H)",
  "Tcell-Endo (L-H)", "Tcell-Endo (H-H)", "Tcell-Endo (H-L)",
  "Endo-Tumor (L-H)", "Endo-Tumor (H-H)", "Endo-Tumor (H-L)",
  "Tcell-Tcell (L-H)", "Tcell-Tcell (H-H)",
  "Tcell-Tumor (L-H)", "Tcell-Tumor (H-H)", "Tcell-Tumor (H-L)",
  "Tumor-Tumor (L-H)", "Tumor-Tumor (H-H)"
)

prox_clust$pair <- factor(prox_clust$pair, levels = newlevels)

##  Pairwise statistical test
stat.test <- prox_clust %>%
  group_by(pair) %>%
  wilcox_test(proximity ~ MFP) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
  add_xy_position(x = "pair", group = "MFP")

##  Factor levels
prox_clust$pair <- factor(prox_clust$pair, levels = unique(test_subtypes$pair))

## Plot
bxp_prox_clust <- ggpubr::ggboxplot(
  prox_clust,
  x = "pair", y = "proximity",
  fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
  add = "jitter", add.params = list(size = 0.4, jitter = 0.08)
) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 16, color = "black"),
    axis.text.x = element_text(size = 16, color = "black", angle = 45, hjust = 1),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 16, color = "black"),
    legend.position = "top", legend.direction = "horizontal"
  ) +
  labs(y = "Proximity of clusters")

Fig4c <- bxp_prox_clust +
  stat_pvalue_manual(subset(test_subtypes, !p.adj.signif == "ns"),
    label = "p.adj.signif", label.size = 5, tip.length = 0.01
  )

##  Combine first two panels
pt_1 <- ggarrange(Fig4a,
  Fig4b,
  nrow = 1, ncol = 2, align = "v", widths = c(1, 1),
  common.legend = TRUE
)

##  Combine last panel
ggarrange(pt_1,
  Fig4c,
  nrow = 2, ncol = 1, align = "v", widths = c(0.5, 1),
  legend = "none"
)
```

## Figure 5

Two slides classified as IE subtypes
Two slides classified as F subtypes which were predicted to have similar cell type abundance

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 22, fig.height = 22}
cell_prob$sample_submitter_id <- cell_prob$Section <- cell_prob$tile_ID <- NULL

##  Combine cell types
combined_cell_prob <- cell_prob %>%
  dplyr::group_by(slide_submitter_id, Coord_X, Coord_Y, celltype, MFP) %>%
  dplyr::summarise(prob_logic = prob > 0.5) %>%
  dplyr::mutate(celltype = ifelse(prob_logic, celltype, "other"))

##  Change cell-type names
combined_cell_prob <- combined_cell_prob %>%
  mutate(celltype = gsub("T_cells", "Tcell", celltype))
combined_cell_prob <- combined_cell_prob %>%
  mutate(celltype = gsub("endothelial_cells", "Endo", celltype))
combined_cell_prob <- combined_cell_prob %>%
  mutate(celltype = gsub("tumor_purity", "Tumor", celltype))
combined_cell_prob <- combined_cell_prob %>%
  mutate(celltype = gsub("CAFs", "CAF", celltype))

combined_cell_prob$TCGA_patient_ID <- substr(combined_cell_prob$slide_submitter_id, 1, 12)

combined_cell_prob <- combined_cell_prob %>%
  dplyr::group_by(slide_submitter_id, Coord_X, Coord_Y) %>%
  dplyr::summarise(celltype_combo = celltype[prob_logic]) %>%
  dplyr::group_by(slide_submitter_id, Coord_X, Coord_Y) %>%
  dplyr::summarise(celltype_combo = paste(celltype_combo, collapse = "&"))

##  Two IE slides ---
IE_A <- "TCGA-BF-AAP4-01A-01-TS1" #  scattered
IE_B <- "TCGA-EE-A2MD-06A-01-TSA" #  confined

## Two FF ---
F_A <- "TCGA-D3-A1Q8-06A-01-TSA" # 0.53 0.53 0.53 0.47
F_B <- "TCGA-ER-A19D-06A-01-TSA" # 0.56 0.51 0.51 0.40

colors <- c(
  "darkgreen", "darkblue", "darkorange", "salmon", "darkred",
  "goldenrod", "gray"
)

names(colors) <- c(
  "CAF", "Tcell", "CAF&Tcell", "Tcell&Tumor", "Tumor",
  "CAF&Tcell&Endo", "other"
)

sel_celltypes <- c("Tcell", "CAF&Tcell", "CAF", "Tcell&Tumor", "CAF&Tcell&Endo", "Tumor")

combined_cell_prob <- combined_cell_prob %>%
  dplyr::mutate(celltype_combo_plot = ifelse(celltype_combo %in% sel_celltypes,
    celltype_combo, "other"
  ))

examples <- c(IE_A, IE_B, F_A, F_B)
gg_maps <- lapply(examples, function(X) {
  example <- cell_prob %>% filter(TCGA_patient_ID %in% substr(X, 1, 12))

  # split data for each "facet"
  df <- split(example, f = example$celltype)

  plot_CAFs <- df$CAFs %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Greens", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    facet_grid(. ~ slide_submitter_id) +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_endo <- df$endothelial_cells %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Purples", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_tcells <- df$T_cells %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Blues", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_tumor <- df$tumor_purity %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_distiller(palette = "Reds", direction = 1, name = "Prob", limits = c(0, 1)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1)

  plot_combined <- combined_cell_prob %>%
    filter(slide_submitter_id %in% X) %>%
    ggplot(aes(x = Coord_X, y = -Coord_Y, fill = celltype_combo_plot)) +
    # geom_point(size = 2.5, shape=15) +
    geom_tile(color = "white") +
    scale_fill_manual(values = colors, labels = names(colors)) +
    theme_void() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      legend.title = element_text(size = 16, color = "black"),
      legend.text = element_text(size = 16, color = "black")
    ) +
    coord_fixed(1) +
    labs(fill = "Assigned cell type")

  if (X == examples[4]) {
    ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor, plot_combined,
      nrow = 5, ncol = 1, align = "v"
    )
  } else {
    ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor, plot_combined,
      nrow = 5, ncol = 1, align = "v", legend = "none"
    )
  }
})

ggarrange(
  plotlist = gg_maps,
  nrow = 1, ncol = 4, align = "h", common.legend = FALSE
)
```

## Figure 6

Hazard ratio (HR) estimate calculated based on univariate cox regression analysis of TCGA overall survival (n=320; deceased = 153) showing the estimated HR (dots) and the 95% confidence interval (bars). Positive (red) and negative (green) HR means that the feature is associated with decreased and increased survival respectively. 

Corresponding receiver operating characteristic (ROC) curve for prediction of 1-year vital status in the CPTAC melanoma cohort (n=36 patients: deceased = 4, living = 32). 

Bar plots showing the elastic-net logistic regression model coefficients for 13 recurrent features (non-zero in >50% of the bootstrap runs). 

Kaplan-Meier (KM) curves displaying the estimated survival probability for two different groups (low and high) defined according to the values of corresponding spatial features. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 22, fig.height = 12}
#  Univariate Cox Regression (Overall Survival)  --------------------------------

survival_liu <- survival_tcga_liu %>%
  dplyr::filter(type == "SKCM", bcr_patient_barcode %in% rownames(tcga_mat)) %>%
  dplyr::select(bcr_patient_barcode, OS, OS.time, ajcc_pathologic_tumor_stage) %>%
  dplyr::rename(
    "pat_id" = "bcr_patient_barcode",
    "tumor_stage" = "ajcc_pathologic_tumor_stage"
  ) %>%
  dplyr::mutate(
    OS = ifelse(OS == "#N/D", NA, OS),
    OS.time = ifelse(OS.time == "#N/D", NA, OS.time),
    tumor_stage = ifelse(tumor_stage == "[Not Available]", NA, tumor_stage)
  ) %>%
  dplyr::mutate(
    OS = as.numeric(OS),
    OS.time = as.numeric(OS.time)
  )
# dplyr::filter(tumor_stage != "Stage IV")

tcga_mat$pat_id <- rownames(tcga_mat)

tcga_survival_mat <- dplyr::inner_join(tcga_mat, survival_liu, by = "pat_id")
tcga_survival_mat <- tcga_survival_mat %>%
  dplyr::filter(!(is.na(OS) & is.na(OS.time)))

tcga_survival_mat <- tcga_survival_mat %>%
  dplyr::filter(!is.na(OS.time) & !is.na(tumor_stage))

cox_features_df <- do.call(rbind, lapply(colnames(tcga_survival_mat)[1:96], function(X) {
  cox_fit <- coxph(Surv(OS.time, OS) ~ get(X), data = tcga_survival_mat) # + tumor_stage
  cox_df <- cox_fit %>%
    tidy() %>%
    dplyr::mutate(term = ifelse(term == "get(X)", X, term))
  cox_df$C <- summary(cox_fit)$concordance[1]
  cox_df$C_se <- summary(cox_fit)$concordance[2]
  return(cox_df)
}))

##  compute HR
cox_features_df <- cox_features_df %>%
  mutate(
    upper = estimate + 1.96 * std.error,
    lower = estimate - 1.96 * std.error
  ) %>%
  mutate(across(all_of(c("estimate", "lower", "upper")), exp))

##  sort features by HR
cox_features_df <- cox_features_df %>%
  filter(term %in% colnames(tcga_survival_mat)[1:96]) %>%
  group_by(term) %>%
  arrange(desc(estimate)) %>%
  dplyr::rename("HR" = "estimate")

## Factor levels
cox_features_df$term <- factor(cox_features_df$term, levels = cox_features_df$term)

## saturate features with very high HR
cox_features_df <- cox_features_df %>%
  mutate(
    HR = ifelse(HR > 5, 5, HR),
    upper = ifelse(upper > 5, 5, upper)
  )

## Subset: Features that HR CI do not cross 1)
cox_features_df_filtered <- cox_features_df %>%
  dplyr::filter((HR > 1 & round(lower, 3) > 1) | (HR < 1 & round(upper, 3) < 1)) %>%
  dplyr::arrange(desc(HR))

## Factor levels
cox_features_df_filtered$term <- factor(cox_features_df_filtered$term, levels = cox_features_df_filtered$term)

OS_HR_filtered <- cox_features_df_filtered %>%
  ggplot(aes(HR, term, color = HR > 1)) +
  geom_vline(xintercept = 1, color = "gray75") +
  geom_linerange(aes(xmin = lower, xmax = upper), size = 1.5, alpha = 0.5) +
  geom_point(size = 2) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("green4", "red3"), guide = "none") +
  xlim(c(0, 5)) +
  labs(y = NULL, x = "OS - HR estimate (95% CI)") +
  theme(
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
    axis.title = element_text(size = 12, color = "black")
  )

#  ROC curve  --------------------------------

##  boostrap learned model on TCGA
load(params$bootstrap_model)

## make predictions on the CPTAC test set for each model
test_pred <- lapply(boot_models, function(x) {
  glmnet::predict.glmnet(x, newx = as.matrix(cptac_mat), type = "response")
})
test_pred.df <- do.call(cbind, test_pred)
test_pred_median <- apply(test_pred.df, 1, median)

## ROC curve performance
###  Median
roc_res <- glmnet::roc.glmnet(test_pred_median, newy = CPTAC_1yearsurv)

### Bootstrap

ROC_curve <- ggplot(roc_res, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
    axis.title = element_text(size = 12, color = "black")
  )

## Features coefficients ----

# retrieve robust coefficients (selected in more that 50% of the bootstrap runs)
nonzero_coeff <- lapply(boot_models, function(x) {
  model_coef <- coef(x)
  return(model_coef@Dimnames[[1]][model_coef@i + 1])
})
nonzero_coeff_sorted <- sort(table(unlist(nonzero_coeff)), decreasing = T)
recurrent_features <- setdiff(names(which(nonzero_coeff_sorted > 50)), "(Intercept)")

bootstrap_coeff <- lapply(boot_models, function(x) {
  model_coef <- coef(x)
  nonZeroFeat <- data.frame(names = model_coef@Dimnames[[1]][model_coef@i + 1], value = model_coef@x)

  ix <- match(recurrent_features, nonZeroFeat$names)
  coeff_value <- nonZeroFeat$value[ix]
  coeff_value[is.na(coeff_value)] <- 0
  return(coeff_value)
})

summary_coeff <- data.frame(
  name = recurrent_features,
  median = apply(do.call(cbind, bootstrap_coeff), 1, median),
  sd = apply(do.call(cbind, bootstrap_coeff), 1, sd)
)


coef_plot <- ggplot(data = summary_coeff, aes(x = reorder(name, -median), y = median)) +
  geom_bar(stat = "identity") +
  labs(x = "Feature", y = "Estimated coefficient") +
  theme_minimal() +
  ylim(c(-2.5, 2.5)) +
  coord_flip() +
  theme(
    axis.text.x = element_text(hjust = 1, size = 12, color = "black", angle = 45),
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.title.y = element_text(size = 12, color = "black"),
    axis.title.x = element_blank()
  )

## Kaplan-Meier Curves of relevant features on the TCGA ------------------------

KM_features_df <- subset(cox_features_df_filtered, term %in% summary_coeff$name)

gg_kms <- lapply(as.character(KM_features_df$term), function(X) {
  # X <- as.character(KM_features_df$term)[1]
  tmp <- tcga_survival_mat %>%
    dplyr::select(c(OS, OS.time, X)) %>%
    dplyr::rename("feature_value" = X) %>%
    dplyr::filter(!is.na(feature_value)) %>%
    dplyr::mutate(inv_rank = rank(-feature_value)) %>%
    dplyr::mutate(category = as.numeric(cut_number(inv_rank, 2)))

  tmp$category <- gsub(1, "High", tmp$category)
  tmp$category <- gsub(2, "Low", tmp$category)
  tmp$category <- factor(tmp$category, levels = c("Low", "High"))

  km_fit_OS <- survfit(Surv(OS.time, OS) ~ category, data = tmp)

  ##  KM plot - 1year
  KM_OS_plot <- ggsurvplot(km_fit_OS,
    palette = c("red3", "green4"),
    # palette = rev(c("red3", "green4")),
    conf.int = TRUE,
    censor = TRUE,
    risk.table = TRUE,
    # xlim = c(0, 360.25),
    # break.x.by = c(60),
    fontsize = 3,
    xlab = "Time in days",
    ggtheme = theme_bw() +
      theme(
        axis.text.y = element_text(hjust = 0, size = 12, color = "black"),
        axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        legend.text = element_text(size = 12)
      )
  )

  KM_OS_plot
})

## 1st row
first_row <- ggarrange(OS_HR_filtered, ROC_curve, coef_plot, nrow = 1, ncol = 3, align = "h")

##  2nd row
second_row <- survminer::arrange_ggsurvplots(gg_kms,
  print = FALSE,
  nrow = 1, ncol = 5,
  risk.table.height = 0.4
)

##  Plot both rows
first_row
second_row
```

## Session Info
```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```