---
title: "Notebook to generate Figure 1d-e and Supplementary Figure 3c: Analysis of Xenium slide with add-on panel"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
author: "Joan Kant"
params: 
    output_dir: "output"
    he_tile_quant: "output/spotlight_rerun/grad_mag10/2_tile_level_quantification/test_tile_predictions_proba.csv"
    bin_size: 1024
    he_alignment_file: "data/ST_melanoma/xenium_skin_panel_w_add_on/Xeniumranger_V1_hSkin_Melanoma_Add_on_FFPE_he_imagealignment.csv"
    sample_id: "xenium_skin_panel_w_add_on"
output:
    html_document:
        code_folding: show
        code_download: true
        theme: cerulean
knit: |
  rmdmatter::renaming_renderer(function(metadata) {
    with(metadata, paste0(stringr::str_replace_all(title, c(" " = "_", "\\:" = "_")), "-",  format(Sys.time(), "%Y%b%d-%H_%M_%S")))})
---
  <style type="text/css">
  .smaller {
    font-size: 10px
    font-style: normal
  }
</style>

## Setup
```{r, notebook-settings}
knitr::opts_knit$set(root.dir = here::here())
options(knitr.table.format = "html", pillar.sigfig = 10)
knitr::opts_chunk$set(fig.height = 8, fig.width = 14, warning = FALSE, message = FALSE, cache = TRUE)

# Unload all previously loaded packages + remove previous environment
pacman::p_load(GaitiLabUtils, glue, data.table, tidyverse, stringr, patchwork, ggplot2)

# Set working directory
set_wd()

# Color palette for cell types of interest
caf_color <- RColorBrewer::brewer.pal(n = 9, name = "Greens")[5]
endo_color <- RColorBrewer::brewer.pal(n = 9, name = "Purples")[5]
tcell_color <- RColorBrewer::brewer.pal(n = 9, name = "Blues")[5]
tumor_color <- RColorBrewer::brewer.pal(n = 9, name = "Reds")[5]
color_celltypes_oi <- sort(setNames(c(caf_color, endo_color, tcell_color, tumor_color, "lightgrey"), c("CAFs", "Endothelial cells", "T cells", "Tumor cells", "Undetermined")))
color_palettes_celltypes_oi <- setNames(c("Greens", "Purples", "Blues", "Reds"), c("CAFs", "Endothelial cells", "T cells", "Tumor cells"))
```

## Load predictions from SPoTLIghT

```{r, load-spotlight-predictions}
he_tile_quant <- data.table::fread(file = file.path(params$he_tile_quant), sep = "\t") %>%
    select(slide_submitter_id, CAFs, T_cells, endothelial_cells, tumor_purity, Coord_X, Coord_Y) %>%
    pivot_longer(
        cols = !c(slide_submitter_id, Coord_X, Coord_Y),
        names_to = "celltype", values_to = "prob"
    ) %>%
    mutate(
        celltype = case_when(
            celltype == "T_cells" ~ "T cells",
            celltype == "endothelial_cells" ~ "Endothelial cells",
            celltype == "tumor_purity" ~ "Tumor cells",
            .default = celltype
        ),
        slide_submitter_id = case_when(
            slide_submitter_id == "xenium-skin-panel" ~ "xenium_skin_panel",
            slide_submitter_id == "xenium-skin-panel-addon" ~ "xenium_skin_panel_w_add_on",
            .default = slide_submitter_id
        ),
    )
```

```{r, load-xenium-skin-panel}
output_dir <- file.path(params$output_dir, params$sample_id)
create_dir(output_dir)

# Load metadata of Xenium sample with annotations
# Convert um to pixels (https://kb.10xgenomics.com/hc/en-us/articles/11636252598925-What-are-the-Xenium-image-scale-factors)
image_scale_factor <- .2125
xenium_meta <- readRDS(glue("data/processed/{params$sample_id}/BANKSY/{params$sample_id}__BANKSY_annotated__metadata.rds")) %>%
    mutate(
        # Relabel to match with cell type labels in spotlight quantification
        Label = factor(ifelse(Label == "Tumor", "Tumor cells", Label)),
        new_x_pixels = x_centroid / image_scale_factor, new_y_pixels = y_centroid / image_scale_factor
    ) %>%
    tibble::rowid_to_column(., "cell_ID")
```

### Aligning the H&E slides (SPoTLIghT predictions)
```{r, align-he-slide-skin-panel}
# Load the image alignment file provided by 10x for the H&E slide
imagealignment <- read.table(params$he_alignment_file, sep = ",") %>%
    data.matrix()

# Select the data for the current slide
he_tile_quant_indiv_sample <- he_tile_quant %>%
    dplyr::filter(slide_submitter_id == params$sample_id)

# Create matrix with X, Y and Z coordinates, setting Z to 0
coord_mat <- he_tile_quant_indiv_sample %>%
    # Add dummy for z-stack
    mutate(Coord_Z = 1) %>%
    select(Coord_X, Coord_Y, Coord_Z) %>%
    data.matrix()

# Apply affine transformation matrix (multiplication of coordinates x image alignment matrix)
coord_mat_transformed <- coord_mat %*% t(imagealignment)
# Add the new coordinates to the spotlight predictions
he_tile_quant_indiv_sample <- he_tile_quant_indiv_sample %>%
    mutate(
        new_x = (coord_mat_transformed[, 1]),
        new_y = (coord_mat_transformed[, 2])
    ) %>%
    tibble::rowid_to_column(., "tile_ID")
```

### Binning cells and tiles

Due to alignment, the tile size, used when tiling the H&E in SPoTLIghT, is lost. Therefore, do binning for both the manually annotated Xenium slide (based on expression) and the new coordinates of the tiles from SPoTLIghT. 

```{r, binning-skin-panel}
# Get extremes for creating a grid that covers all cells and tiles
max_x <- plyr::round_any(max(c(xenium_meta$new_x_pixels, he_tile_quant_indiv_sample$new_x)), params$bin_size * 2, ceiling)
max_y <- plyr::round_any(max(c(xenium_meta$new_y_pixels, he_tile_quant_indiv_sample$new_y)), params$bin_size * 2, ceiling)
min_x <- plyr::round_any(min(c(xenium_meta$new_x_pixels, he_tile_quant_indiv_sample$new_x)), params$bin_size * 2, floor)
min_y <- plyr::round_any(min(c(xenium_meta$new_y_pixels, he_tile_quant_indiv_sample$new_y)), params$bin_size * 2, floor)

# Create bins
x_bins <- data.frame(bin_xcoord = seq(min_x, plyr::round_any(max_x, params$bin_size, ceiling),
    by = params$bin_size
)) %>% tibble::rowid_to_column(., "bin_ID")
y_bins <- data.frame(bin_ycoord = seq(min_y, plyr::round_any(max_y, params$bin_size, ceiling),
    by = params$bin_size
)) %>% tibble::rowid_to_column(., "bin_ID")

print(range(x_bins$bin_xcoord))
print(range(y_bins$bin_ycoord))

# Create full grid
bin_grid <- expand.grid(x_bins$bin_xcoord, y_bins$bin_ycoord) %>%
    data.frame() %>%
    rename(bin_xcoord = Var1, bin_ycoord = Var2)

ncells_xenium_binxy <- data.frame(
    bin_x = findInterval(xenium_meta$new_x_pixels, x_bins$bin_xcoord),
    bin_y = findInterval(xenium_meta$new_y_pixels, y_bins$bin_ycoord)
) %>% tibble::rowid_to_column(., "cell_ID")

ntiles_binxy <- data.frame(
    bin_x = findInterval(he_tile_quant_indiv_sample$new_x, x_bins$bin_xcoord),
    bin_y = findInterval(he_tile_quant_indiv_sample$new_y, y_bins$bin_ycoord)
) %>% tibble::rowid_to_column(., "tile_ID")

# # Add the binx, biny
xenium_meta <- xenium_meta %>%
    # Add assigned bin to dataframe
    left_join(ncells_xenium_binxy, by = "cell_ID") %>%
    # Assign the actual coords of the bin
    left_join(x_bins, by = c(bin_x = "bin_ID")) %>%
    left_join(y_bins, by = c(bin_y = "bin_ID"))

he_tile_quant_indiv_sample <- he_tile_quant_indiv_sample %>%
    # Add assigned bin to dataframe
    left_join(ntiles_binxy, by = "tile_ID") %>%
    # Assign the actual coords of the bin
    left_join(x_bins, by = c(bin_x = "bin_ID")) %>%
    left_join(y_bins, by = c(bin_y = "bin_ID"))
```


### Figure 1d and Supplementary Figure 3c - Binned Spatial Maps

Row 1 = Xenium and row 2 = SPoTLIghT

```{r, spatial-maps-skin-panel, fig.width = 20, fig.height = 10}
p_xenium_binned <- lapply(
    sort(names(color_celltypes_oi)[1:4]),
    function(current_celltype) {
        xenium_by_bin <- xenium_meta %>%
            filter(Label == current_celltype) %>%
            # Grouping by bin
            group_by(bin_xcoord, bin_ycoord) %>%
            count()

        p_xenium_binned <- ggplot(data = xenium_by_bin, aes(x = bin_xcoord, y = bin_ycoord, fill = n)) +
            geom_tile() +
            labs(
                title = current_celltype,
                subtitle = glue::glue("No. cells={formatC(sum(xenium_by_bin$n), big.mark = ',')}")
            ) +
            scale_fill_distiller(
                palette = color_palettes_celltypes_oi[current_celltype],
                direction = 1, name = "Cells",
                guide = guide_colorbar(
                    barwidth = unit(9, "lines"),
                    barheight = unit(.9, "lines"), title.position = "top"
                ), limits = c(0, NA)
            )
        return(p_xenium_binned)
    }
)

p_he_binned_prob <- lapply(
    sort(names(color_celltypes_oi)[1:4]),
    function(current_celltype) {
        he_grouped_by_bin <- he_tile_quant_indiv_sample %>%
            filter(celltype == current_celltype) %>%
            group_by(bin_xcoord, bin_ycoord) %>%
            summarise(
                n = n(),
                mean_prob = mean(prob)
            )

        colorbar_min <- plyr::round_any(quantile(he_grouped_by_bin$mean_prob, .05), .05, floor)
        colorbar_max <- plyr::round_any(quantile(he_grouped_by_bin$mean_prob, .95), .05, ceiling)

        he_grouped_by_bin <- he_grouped_by_bin %>% mutate(mean_prob = case_when(
            mean_prob > colorbar_max ~ colorbar_max,
            mean_prob < colorbar_min ~ colorbar_min,
            .default = mean_prob
        ))

        p_he_binned <- ggplot(data = he_grouped_by_bin, aes(x = bin_xcoord, y = bin_ycoord, fill = mean_prob)) +
            geom_tile() +
            labs(
                title = current_celltype,
                subtitle = glue::glue("No. tiles={formatC(sum(he_grouped_by_bin$n), big.mark = ',')}")
            ) +
            scale_fill_distiller(
                palette = color_palettes_celltypes_oi[current_celltype],
                direction = 1, name = "Prob",
                guide = guide_colorbar(
                    barwidth = unit(9, "lines"),
                    barheight = unit(.9, "lines"), title.position = "top"
                ),
                limits = c(colorbar_min, colorbar_max)
            )

        return(p_he_binned)
    }
)

p_combined_proba <- wrap_plots(c(p_xenium_binned, p_he_binned_prob)) + plot_layout(nrow = 2) +
    plot_annotation(subtitle = glue("binsize={params$bin_size}px"))

p_combined_proba <- p_combined_proba &
    GaitiLabUtils::default_theme() &
    xlim(c(min_x - params$bin_size, max_x + params$bin_size)) &
    ylim(c(min_y - params$bin_size, max_y + params$bin_size)) &
    labs(x = "X (px)", y = "Y (px)") & theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom", aspect.ratio = 1,
    axis.title = element_text(face = "plain")
)

p_combined_proba
ggsave(p_combined_proba, filename = glue("{output_dir}/spatial_maps_proba.pdf"), width = 20, height = 10)
```

### Figure 1e - Comparison Xenium spatial cell type distribution vs SPoTLIghT's predictions

Compare number of cells (Xenium) vs mean probability (SPoTLIghT) per bin.

Of note, each dot in the correlation plot is a bin. 

> Bins that do not have any cells (Xenium) AND that also do not have any tiles (SPoTLIghT) are removed. 

```{r, quant-xenium-vs-spotlight, fig.width = 20, fig.height = 5}
p_scatter <- lapply(sort(names(color_celltypes_oi)[1:4]), function(current_celltype) {
    # Comparison: Xenium vs Spotlight (probabilities)
    xenium_by_bin <- xenium_meta %>%
        filter(Label == current_celltype) %>%
        # Grouping by bin
        group_by(bin_xcoord, bin_ycoord) %>%
        summarise(ncells_xenium = n())

    he_grouped_by_bin <- he_tile_quant_indiv_sample %>%
        filter(celltype == current_celltype) %>%
        group_by(bin_xcoord, bin_ycoord) %>%
        summarise(
            ntiles_spotlight = n(),
            mean_prob_spotlight = mean(prob)
        )

    combined_df_w_prob <- merge(xenium_by_bin, he_grouped_by_bin, by = c("bin_xcoord", "bin_ycoord"), all = TRUE)
    combined_df_w_prob[is.na(combined_df_w_prob)] <- 0
    head(combined_df_w_prob)
    combined_df_w_prob <- combined_df_w_prob %>% filter(ncells_xenium != 0, ntiles_spotlight != 0)

    p <- ggpubr::ggscatter(combined_df_w_prob,
        x = "ncells_xenium", y = "mean_prob_spotlight",
        size = 1, fill = NA,
        add = "reg.line", # Add regressin line
        add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
        conf.int = TRUE # Add confidence interval
    ) +
        # Add correlation coefficient
        ggpubr::stat_cor(method = "spearman") +
        labs(
            title = current_celltype,
            subtitle = glue("No. bins={formatC(nrow(combined_df_w_prob), big.mark = ',')}")
        ) +
        scale_y_continuous(breaks = scales::pretty_breaks()) +
        scale_x_continuous(breaks = scales::pretty_breaks())
    return(p)
})

p_combined <- wrap_plots(p_scatter) + plot_layout(nrow = 1, axis_titles = "collect") & GaitiLabUtils::default_theme() &
    labs(
        x = "No. cells from Xenium annotation",
        y = "Mean predicted prob by SPoTLIghT",
    ) & theme(aspect.ratio = 1, axis.title = element_text(face = "plain"))
p_combined

ggsave(p_combined, filename = glue("{output_dir}/{params$sample_id}__binsize{params$bin_size}__corr__proba.pdf"), width = 20, height = 5)
```

## Session Info
```{r, session-info, echo = FALSE}
sessionInfo()
```

