---
title: "Notebook to generate Figure 2" 
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
author: 
    - Óscar Lapuente-Santana
    - Joan Kant
params: 
    spatial_features_dir: "data/features_tcga_skcm_FF"
    tcga_bagaev_subtypes: "data/bagaev_file.xlsx"
output:
    html_document:
        code_folding: hide
        code_download: true
knit: |
  rmdmatter::renaming_renderer(function(metadata) {
    with(metadata, paste0(stringr::str_replace_all(title, " ", "_"), "-",  format(Sys.time(), "%Y%b%d-%H_%M_%S")))})
---
  <style type="text/css">
  .smaller {
    font-size: 10px
    font-style: normal
  }
</style>

## Setup 
```{r, notebook-settings, warning=FALSE, message=FALSE}
knitr::opts_knit$set(root.dir = here::here())
options(knitr.table.format = "html", pillar.sigfig = 10)
# change default settings to fit plots
knitr::opts_chunk$set(fig.height = 8, fig.width = 14, warning = FALSE, message = FALSE, cache = TRUE)

pacman::p_load(GaitiLabUtils, data.table, tidyverse, stringr, ggplot2, readxl, ggpubr, rstatix)

# Set working directory
set_wd()

# PLEASE UNCOMMENT WHEN WANTING TO KNIT
# params <- list()
# params$spatial_features_dir <- "data/features_tcga_skcm_FF"
# params$tcga_bagaev_subtypes <- "data/bagaev_file.xlsx"
# packages
# pacman::p_load(dplyr, corrplot, )
# suppressPackageStartupMessages({
#     library("dplyr")
#     library("corrplot")
#     library("gplots")
#     library("RColorBrewer")
#     library("Matrix")
#     library("ggplot2")
#     library("ggforce")
#     library("ggpubr")
#     library("reshape2")
#     library("ggrepel")
#     library("tidyr")
#     library("plyr")
#     library("purrr")
#     library("gridExtra")
#     library("rstatix")
#     library("ComplexHeatmap")
#     library("survminer")
#     library("survival")
#     library("ggfortify")
# })

colors_mfp <- c("#64a860", "#9970c1", "#cc545e", "#b98d3e")
names(colors_mfp) <- c("D", "F", "IE/F", "IE")
```


```{r, prep-data}
bagaev_mfp <- readxl::read_xlsx(params$tcga_bagaev_subtypes, sheet = 1, skip = 1)
bagaev_mfp <- bagaev_mfp %>%
    dplyr::filter(Cohort == "TCGA-SKCM") %>%
    rename(TCGA_patient_ID = Sample)

##  Read cell type probabilities
cell_prob <- data.table::fread(file.path(params$spatial_features_dir, "tcga_validation_tile_predictions_proba.csv"), sep = "\t")

## Pivot to long df
cell_prob <- cell_prob %>%
    dplyr::select(slide_submitter_id, CAFs, T_cells, endothelial_cells, tumor_purity, Coord_X, Coord_Y) %>%
    pivot_longer(
        cols = !c(slide_submitter_id, Coord_X, Coord_Y),
        names_to = "celltype", values_to = "prob"
    )

##  Add bagaev subtypes data
tmp <- bagaev_mfp %>% dplyr::select(TCGA_patient_ID, MFP)
cell_prob$TCGA_patient_ID <- substr(cell_prob$slide_submitter_id, 1, 12)
cell_prob <- inner_join(cell_prob, tmp, by = "TCGA_patient_ID")
```

## Fig 2a

Cell type-specific probability maps for four examples of melanoma slides from distinct microenvironments: immune-depleted (D), fibrotic (F), immune-enriched fibrotic (IE/F) and immune-enriched (IE) subtypes. 

** Note that the full slide is shown here. **

```{r, vis-data, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 22, fig.height = 22}
##  Desert
D_example <- "TCGA-YG-AA3P"
# TCGA-Z2-AA3S
# TCGA-D3-A8GL

##  Fibrotic
F_example <- "TCGA-EE-A3JB"
# TCGA-D3-A2JA
#  TCGA-FR-A2OS

##  Immune-enrinched fibrotic
IEF_example <- "TCGA-EE-A2ME" # IE/F
# TCGA-FS-A1ZH
# TCGA-D3-A1Q3

##  Immune-enrinched
IE_example <- "TCGA-ER-A1A1"
# TCGA-D3-A8GM
# TCGA-WE-A8K4

##  Visualize selected slides ##
examples <- c(D_example, F_example, IEF_example, IE_example)

##  Desert
D_example <- "TCGA-YG-AA3P"
# TCGA-Z2-AA3S
# TCGA-D3-A8GL

##  Fibrotic
F_example <- "TCGA-EE-A3JB"
# TCGA-D3-A2JA
#  TCGA-FR-A2OS

##  Immune-enrinched fibrotic
IEF_example <- "TCGA-EE-A2ME" # IE/F
# TCGA-FS-A1ZH
# TCGA-D3-A1Q3

##  Immune-enrinched
IE_example <- "TCGA-ER-A1A1"
# TCGA-D3-A8GM
# TCGA-WE-A8K4

##  Visualize selected slides ##
examples <- c(D_example, F_example, IEF_example, IE_example)

gg_maps <- lapply(examples, function(X) {
    example <- cell_prob %>% filter(TCGA_patient_ID %in% X)

    # split data for each "facet"
    df <- split(example, f = example$celltype)

    plot_CAFs <- df$CAFs %>%
        ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
        geom_tile(color = "white") +
        scale_fill_distiller(palette = "Greens", direction = 1, name = "Prob", limits = c(0, 1)) +
        theme_void() +
        theme(
            axis.title = element_blank(),
            axis.text = element_blank(),
            legend.title = element_text(size = 16, color = "black"),
            legend.text = element_text(size = 16, color = "black")
        ) +
        coord_fixed(1)

    plot_endo <- df$endothelial_cells %>%
        ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
        # geom_point(size = 2.5, shape=15) +
        geom_tile(color = "white") +
        scale_fill_distiller(palette = "Purples", direction = 1, name = "Prob", limits = c(0, 1)) +
        theme_void() +
        theme(
            axis.title = element_blank(),
            axis.text = element_blank(),
            legend.title = element_text(size = 16, color = "black"),
            legend.text = element_text(size = 16, color = "black")
        ) +
        coord_fixed(1)

    plot_tcells <- df$T_cells %>%
        ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
        # geom_point(size = 2.5, shape=15) +
        geom_tile(color = "white") +
        scale_fill_distiller(palette = "Blues", direction = 1, name = "Prob", limits = c(0, 1)) +
        theme_void() +
        theme(
            axis.title = element_blank(),
            axis.text = element_blank(),
            legend.title = element_text(size = 16, color = "black"),
            legend.text = element_text(size = 16, color = "black")
        ) +
        coord_fixed(1)

    plot_tumor <- df$tumor_purity %>%
        ggplot(aes(x = Coord_X, y = -Coord_Y, fill = prob)) +
        # geom_point(size = 2.5, shape=15) +
        geom_tile(color = "white") +
        scale_fill_distiller(palette = "Reds", direction = 1, name = "Prob", limits = c(0, 1)) +
        theme_void() +
        theme(
            axis.title = element_blank(),
            axis.text = element_blank(),
            legend.title = element_text(size = 16, color = "black"),
            legend.text = element_text(size = 16, color = "black")
        ) +
        coord_fixed(1)

    if (X == examples[4]) {
        ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor,
            nrow = 4, ncol = 1, align = "v"
        )
    } else {
        ggarrange(plot_CAFs, plot_endo, plot_tcells, plot_tumor,
            nrow = 4, ncol = 1, align = "v", legend = "none"
        )
    }
})

ggarrange(
    plot_list = gg_maps,
    align = "h", common.legend = TRUE
)

# ggpubr::ggarrange(
#     plot_list = gg_maps,
#     ncol = 4, align = "h", common.legend = TRUE
# )
```

## Fig 2b and c

Slide-level quantification of each cell type
Co-localization assessment of cell type pairs across the distinct microenvironment subtypes 

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width = 12, fig.height = 12}
# Cell type probability -------------------------------------------

## Calculate median probability for each slide
cell_prob_mean <- cell_prob %>%
    dplyr::group_by(TCGA_patient_ID, celltype, MFP) %>%
    dplyr::summarise(mean_prob = mean(prob))

cell_prob_mean$MFP <- factor(cell_prob_mean$MFP, levels = names(colors_mfp))
cell_prob_mean <- cell_prob_mean %>% ungroup()

##  Pairwise statistical test #
stat.test <- cell_prob_mean %>%
    group_by(celltype) %>%
    wilcox_test(mean_prob ~ MFP) %>%
    adjust_pvalue(method = "BH") %>%
    add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
    add_xy_position(x = "celltype", group = "MFP") %>%
    filter(!p.adj.signif == "ns")

##  Add factor levels
cell_prob_mean$celltype <- factor(cell_prob_mean$celltype,
    levels = c(
        "CAFs", "endothelial_cells",
        "T_cells", "tumor_purity"
    )
)

## Plot
bxp_prob <- ggpubr::ggboxplot(
    cell_prob_mean,
    x = "celltype", y = "mean_prob",
    fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
    add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
    scale_x_discrete(labels = c("CAFs", "Endo cells", "T cells", "Tumor cells")) +
    theme_minimal() +
    theme(
        axis.text = element_text(size = 16, color = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16, color = "black"),
        legend.position = "top", legend.direction = "horizontal",
        legend.title = element_text(size = 16, color = "black"),
        legend.text = element_text(size = 16, color = "black")
    ) +
    labs(y = "Mean probability")

fig2a <- bxp_prob +
    stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

# Cell type co-localization -------------------------------------------

cell_coloc <- data.table::fread(
    file = file.path(params$spatial_features_dir, "FF_features_coloc_fraction.csv"),
    sep = "\t"
)

##  Change cell-type names
cell_coloc <- cell_coloc %>%
    mutate(pair = gsub("T_cells", "Tcell", pair))
cell_coloc <- cell_coloc %>%
    mutate(pair = gsub("endothelial_cells", "Endo", pair))
cell_coloc <- cell_coloc %>%
    mutate(pair = gsub("tumor_purity", "Tumor", pair))
cell_coloc <- cell_coloc %>%
    mutate(pair = gsub("CAFs", "CAF", pair))

##  Add Bagaev subtypes
tmp <- bagaev_mfp %>% select(TCGA_patient_ID, MFP)
cell_coloc$TCGA_patient_ID <- substr(cell_coloc$slide_submitter_id, 1, 12)
cell_coloc <- inner_join(cell_coloc, tmp, by = "TCGA_patient_ID")
cell_coloc$MFP <- factor(cell_coloc$MFP, levels = names(colors_mfp))

cell_coloc$pair <- factor(cell_coloc$pair, levels = c(
    "CAF-Endo", "CAF-Tcell", "CAF-Tumor",
    "Endo-Tumor", "Tcell-Endo", "Tcell-Tumor"
))
##  Pairwise statistical test
stat.test <- cell_coloc %>%
    group_by(pair) %>%
    wilcox_test(frac ~ MFP) %>%
    adjust_pvalue(method = "BH") %>%
    add_significance("p.adj")

## Add p-values onto the box plots
test_subtypes <- stat.test %>%
    add_xy_position(x = "pair", group = "MFP") %>%
    filter(!p.adj.signif == "ns")

##  Add factor levels
cell_coloc$pair <- factor(cell_coloc$pair, levels = unique(test_subtypes$pair))

##  Plot
bxp_coloc <- ggpubr::ggboxplot(
    cell_coloc,
    x = "pair", y = "frac",
    fill = "MFP", color = "MFP", alpha = 0.8, palette = colors_mfp,
    add = "jitter", add.params = list(size = 0.8, jitter = 0.08)
) +
    theme_minimal() +
    theme(
        axis.text = element_text(size = 16, angle = 45, hjust = 1, color = "black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16, color = "black"),
        legend.position = "top", legend.direction = "horizontal",
        legend.title = element_text(size = 16, color = "black"),
        legend.text = element_text(size = 16, color = "black")
    ) +
    labs(y = "Co-localization")

fig2b <- bxp_coloc +
    stat_pvalue_manual(test_subtypes, label = "p.adj.signif", label.size = 5, tip.length = 0.01)

ggarrange(fig2a,
    fig2b,
    nrow = 1, ncol = 2, align = "h", common.legend = TRUE
)
```

## Session Info

```{r, session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```