---
title: "Notebook to generate Figure 6: Survival Analysis" 
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
author: 
    - Óscar Lapuente-Santana
    - Federica Eduati
params: 
    spatial_features_dir: "data/features_tcga_skcm_FF"
    tcga_bagaev_subtypes: "data/bagaev_file.xlsx"
    tcga_survival: "data/survival.xlsx"
    bootstrap_model: "data/bootstrap_model.RData"
    spatial_features_matrix_tcga: "data/spatial_features_matrix_TCGA.csv"
    spatial_features_matrix_cptac: "data/spatial_features_matrix_CPTAC.csv"
output:
    html_document:
        code_folding: show
        code_download: true
        theme: cerulean
knit: |
  rmdmatter::renaming_renderer(function(metadata) {
    with(metadata, paste0(stringr::str_replace_all(title, c(" " = "_", "\\:" = "_")), "-",  format(Sys.time(), "%Y%b%d-%H_%M_%S")))})
---
  <style type="text/css">
  .smaller {
    font-size: 10px
    font-style: normal
  }
</style>

## Setup 

```{r, notebook-settings, warning=FALSE, message=FALSE}
knitr::opts_knit$set(root.dir = here::here())
options(knitr.table.format = "html", pillar.sigfig = 10)
# change default settings to fit plots
knitr::opts_chunk$set(fig.height = 8, fig.width = 14, warning = FALSE, message = FALSE, cache = TRUE)

pacman::p_load(GaitiLabUtils, data.table, tidyverse, stringr, ggplot2, ggpubr, survival, glmnet, survminer, rstatix)

# Set working directory
set_wd()
```

## Prepare/format data

```{r, prep-data}
bagaev_mfp <- readxl::read_xlsx(params$tcga_bagaev_subtypes, sheet = 1, skip = 1)
bagaev_mfp <- bagaev_mfp %>%
  dplyr::filter(Cohort == "TCGA-SKCM") %>%
  rename(TCGA_patient_ID = Sample)

# Read Survival data for TCGA data ---------------------------------------------

survival_tcga_liu <- readxl::read_xlsx(path = params$tcga_survival, sheet = 1)

# Read TCGA-SKCM features ------------------------------------------------------
tcga_mat <- data.table::fread(file = params$spatial_features_matrix_tcga) %>%
  tibble::column_to_rownames(var = "V1") %>%
  mutate(pat_id = rownames(.))

# Read CPTAC-CM features ------------------------------------------------------

cptac_mat <- data.table::fread(file = params$spatial_features_matrix_cptac) %>% tibble::column_to_rownames(var = "V1")
```


## Fig. 6a - Univariate Cox Regression (Overall Survival)

Hazard ratio (HR) estimate calculated based on univariate cox regression analysis of TCGA overall survival (n=320; deceased = 153) showing the estimated HR (dots) and the 95% confidence interval (bars). Positive (red) and negative (green) HR means that the feature is associated with decreased and increased survival respectively. 

```{r, uni-cox-regression-tcga, fig.width = 22, fig.height = 12}
#  Univariate Cox Regression (Overall Survival)  --------------------------------

survival_liu <- survival_tcga_liu %>%
  dplyr::filter(type == "SKCM", bcr_patient_barcode %in% rownames(tcga_mat)) %>%
  dplyr::select(bcr_patient_barcode, OS, OS.time, ajcc_pathologic_tumor_stage) %>%
  dplyr::rename(
    "pat_id" = "bcr_patient_barcode",
    "tumor_stage" = "ajcc_pathologic_tumor_stage"
  ) %>%
  dplyr::mutate(
    OS = ifelse(OS == "#N/D", NA, OS),
    OS.time = ifelse(OS.time == "#N/D", NA, OS.time),
    tumor_stage = ifelse(tumor_stage == "[Not Available]", NA, tumor_stage)
  ) %>%
  dplyr::mutate(
    OS = as.numeric(OS),
    OS.time = as.numeric(OS.time)
  )

# Combine matrices/dataframes
tcga_survival_mat <- dplyr::inner_join(tcga_mat, survival_liu, by = "pat_id")
tcga_survival_mat <- tcga_survival_mat %>%
  dplyr::filter(
    # Remove observations where OS and OS.time are both missing
    !(is.na(OS) & is.na(OS.time)),
    # Remove observations where OS.time and tumor stage are missing
    !(is.na(OS.time) & is.na(tumor_stage))
  )

cox_features_df <- do.call(rbind, lapply(colnames(tcga_survival_mat)[1:96], function(X) {
  cox_fit <- coxph(Surv(OS.time, OS) ~ get(X), data = tcga_survival_mat) # + tumor_stage
  cox_df <- cox_fit %>%
    tidy() %>%
    dplyr::mutate(term = ifelse(term == "get(X)", X, term))
  cox_df$C <- summary(cox_fit)$concordance[1]
  cox_df$C_se <- summary(cox_fit)$concordance[2]
  return(cox_df)
})) %>%
  ##  compute HR
  dplyr::mutate(
    upper = estimate + 1.96 * std.error,
    lower = estimate - 1.96 * std.error
  ) %>%
  dplyr::mutate(across(all_of(c("estimate", "lower", "upper")), exp)) %>%
  ##  sort features by HR
  dplyr::filter(term %in% colnames(tcga_survival_mat)[1:96]) %>%
  dplyr::group_by(term) %>%
  dplyr::arrange(desc(estimate)) %>%
  dplyr::rename("HR" = "estimate") %>%
  mutate(term = factor(term, levels = unique(term))) %>%
  ## saturate features with very high HR
  dplyr::mutate(
    HR = ifelse(HR > 5, 5, HR),
    upper = ifelse(upper > 5, 5, upper)
  )

## Subset: Features that HR CI do not cross 1)
cox_features_df_filtered <- cox_features_df %>%
  dplyr::filter((HR > 1 & round(lower, 3) > 1) | (HR < 1 & round(upper, 3) < 1)) %>%
  dplyr::arrange(desc(HR)) %>%
  ## Factor levels
  mutate(term = factor(term, levels = unique(term)))


OS_HR_filtered <- cox_features_df_filtered %>%
  ggplot(aes(HR, term, color = HR > 1)) +
  geom_vline(xintercept = 1, color = "gray75") +
  geom_linerange(aes(xmin = lower, xmax = upper), linewidth = 1.5, alpha = 0.5) +
  geom_point(size = 2) +
  theme_bw(base_size = 12) +
  scale_color_manual(values = c("green4", "red3"), guide = "none") +
  xlim(c(0, 5)) +
  labs(y = NULL, x = "OS - HR estimate (95% CI)") +
  theme(
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
    axis.title = element_text(size = 12, color = "black")
  )

OS_HR_filtered
```

## Fig. 6b - CPTAC Validation Dataset

Corresponding receiver operating characteristic (ROC) curve for prediction of 1-year vital status in the CPTAC melanoma cohort (n=36 patients: deceased = 4, living = 32). 

```{r, prep-cptac-data}
#  ROC curve  --------------------------------
##  boostrap learned model on TCGA
load(params$bootstrap_model)

## make predictions on the CPTAC test set for each model
test_pred <- lapply(boot_models, function(x) {
  glmnet::predict.glmnet(x, newx = as.matrix(cptac_mat), type = "response")
})
test_pred.df <- do.call(cbind, test_pred)
test_pred_median <- apply(test_pred.df, 1, median)

## ROC curve performance
###  Median
roc_res <- glmnet::roc.glmnet(test_pred_median, newy = CPTAC_1yearsurv)

### Bootstrap
ROC_curve <- ggplot(roc_res, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray") +
  labs(x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
    axis.title = element_text(size = 12, color = "black")
  )
ROC_curve
```

## Fig. 6c - Elastic-net logistic regression model coefficients
Bar plots showing the elastic-net logistic regression model coefficients for 13 recurrent features (non-zero in >50% of the bootstrap runs). 

```{r, bar-plots-coeff-features}
## Features coefficients ----

# retrieve robust coefficients (selected in more that 50% of the bootstrap runs)
nonzero_coeff <- lapply(boot_models, function(x) {
  model_coef <- coef(x)
  return(model_coef@Dimnames[[1]][model_coef@i + 1])
})
nonzero_coeff_sorted <- sort(table(unlist(nonzero_coeff)), decreasing = TRUE)
recurrent_features <- setdiff(names(which(nonzero_coeff_sorted > 50)), "(Intercept)")

bootstrap_coeff <- lapply(boot_models, function(x) {
  model_coef <- coef(x)
  nonZeroFeat <- data.frame(names = model_coef@Dimnames[[1]][model_coef@i + 1], value = model_coef@x)

  ix <- match(recurrent_features, nonZeroFeat$names)
  coeff_value <- nonZeroFeat$value[ix]
  coeff_value[is.na(coeff_value)] <- 0
  return(coeff_value)
})

summary_coeff <- data.frame(
  name = recurrent_features,
  median = apply(do.call(cbind, bootstrap_coeff), 1, median),
  sd = apply(do.call(cbind, bootstrap_coeff), 1, sd)
)

coef_plot <- ggplot(data = summary_coeff, aes(x = reorder(name, -median), y = median)) +
  geom_bar(stat = "identity") +
  labs(x = "Feature", y = "Estimated coefficient") +
  theme_minimal() +
  ylim(c(-2.5, 2.5)) +
  coord_flip() +
  theme(
    axis.text.x = element_text(hjust = 1, size = 12, color = "black", angle = 45),
    axis.text.y = element_text(hjust = 1, size = 12, color = "black"),
    axis.title.y = element_text(size = 12, color = "black"),
    axis.title.x = element_blank()
  )
coef_plot
```

## Fig 6d. Kaplan-Meier (KM) curves
Kaplan-Meier (KM) curves displaying the estimated survival probability for two different groups (low and high) defined according to the values of corresponding spatial features.

```{r, vis-kaplan-meier}
## Kaplan-Meier Curves of relevant features on the TCGA ------------------------

KM_features_df <- subset(cox_features_df_filtered, term %in% summary_coeff$name)

gg_kms <- lapply(as.character(KM_features_df$term), function(X) {
  tmp <- tcga_survival_mat %>%
    dplyr::select(c(OS, OS.time, X)) %>%
    dplyr::rename("feature_value" = X) %>%
    dplyr::filter(!is.na(feature_value)) %>%
    dplyr::mutate(inv_rank = rank(-feature_value)) %>%
    dplyr::mutate(category = as.numeric(cut_number(inv_rank, 2)))

  tmp$category <- gsub(1, "High", tmp$category)
  tmp$category <- gsub(2, "Low", tmp$category)
  tmp$category <- factor(tmp$category, levels = c("Low", "High"))

  km_fit_OS <- survival::survfit(survival::Surv(OS.time, OS) ~ category, data = tmp)

  ##  KM plot
  KM_OS_plot <- survminer::ggsurvplot(km_fit_OS,
    data = tmp,
    palette = c("red3", "green4"),
    conf.int = TRUE,
    censor = TRUE,
    risk.table = TRUE,
    fontsize = 3,
    xlab = "Time in days",
    ggtheme = theme_bw() +
      theme(
        axis.text.y = element_text(hjust = 0, size = 12, color = "black"),
        axis.text.x = element_text(hjust = 0.5, size = 12, color = "black"),
        axis.title = element_text(size = 12, color = "black"),
        legend.text = element_text(size = 12)
      )
  )
})
```

## Visualization 
```{r, vis-combine, fig.width = 22, fig.height = 12}
## 1st row
first_row <- ggpubr::ggarrange(OS_HR_filtered, ROC_curve, coef_plot, nrow = 1, ncol = 3, align = "h")

##  2nd row
second_row <- survminer::arrange_ggsurvplots(gg_kms,
  print = FALSE,
  nrow = 1, ncol = 5,
  risk.table.height = 0.4
)

##  Plot both rows
first_row
second_row
```


## Session Info

```{r, session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```